Density-dependent selection and the limits of relative fitness

### 2017-12-13 manuscript.tex
l

\textit{Keywords}: $r$/$K$ selection, eco-evolutionary synthesis.

@@

TBA

@@

There are a variety of different measures of fitness. Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, because different measures  may be more useful in different circumstances. But it should be clear how the measure being used is connected to the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}. While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness seems largely divorced from population biology. It has even been proposed that relative fitness be justified from more abstract measure-theoretical arguments, abandoning population biology altogether \citep{wagner_2010}.

In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate. In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$. Here we can rescale $W$ however we please and replace it with ``relative fitness'' $w$ without affecting the ratio $\frac{W_i}{\overline{W}}=\frac{w_i}{\overline{w}}$. In continuous time, the canonical selection equation is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W$ is replaced by the intrinsic exponential growth rate $r$ \citep[pp. 26]{crow_1970}. If there are two types present, a wildtype and a mutant for instance, then the continuous time canonical selection equation can be written as
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where the constant selection coefficient $s$ is the difference in $r$ between types. The corresponding adaptive sweeps follow a logistic curve.

@@

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. Adaptive dynamics currently provides a powerful framework for addressing the complex feedbacks between evolutionary change and population density \citep{diekmann_2004}. However, the focus of adaptive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''. We emphasize that our focus here is the description of the time-dependent process of selection itself, which is particularly critical for making sense of evolution at the genetic level for which we now have abundant data.

Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. Rather attempting to make sense of relative fitness in the standard models of population growth mentioned above (e.g. \cite{mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher, constant-$N$, relative fitness model. Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}. Like the Wright-Fisher model, the classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}. Unlike the Wright-Fisher model, generations can overlap in the lottery model. We generalize the lottery model to allow population density to vary, and show that this model can be interpreted as a density-dependent generalization of the Wright-Fisher model with overlapping generations.

We show that when this model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges. Futhermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded populations. In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities. This is largely an artifact of the models historically used in the density-dependent selection literature, which ignore relative contests.

Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ). We then...

@@

When multiple propagules land on the same territory, the victor is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$. type $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these. Ignoring fluctuations about these two expectations (due to our no-drift, large $T$, large $n_i$ approximation), type $i$'s territorial acquisition is given by
\begin{equation}
\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G), \label{eq:growthsumuncoupled}
\end{equation}
in our extended lottery model, where the sum only includes territories with at least one propagule present.

@@

Eq. \eqref{eq:lottery} breaks down for types with low propagule density ($l_i\ll 1$) because territorial acquisition is then not correctly represented by a lottery in each territory with the mean propagule density. Instead, a rare type's propagules only make it to a few territories where at least one of its propagule present. In our extension of the classic lottery model, we correct (Eq.~\ref{eq:delttot}) to account for this.

@@

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ type is a rare invader in a high density population, determining its invasion fitness \citep{metz_1992}. The third term, $A_i c_i/\overline{c}$, represents competitive victories when the $i$ type is abundant. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $m_i/M$. If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$). Note that not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied; total population density thus obeys 
\begin{equation}
\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
\end{equation}

@@

We are now in a position to analyze the validity of Eq.~\eqref{eq:canonical} more explicitly. In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's general argument \citep{macarthur_1967} for the $r$/$K$ scheme never precluded this possibility. Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth. To understand why this poses a problem for Eq.~\eqref{eq:canonical}, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in type $i$'s mortality rate due to crowding (for simplicity, there are no deaths when uncrowded). Then, starting from a monomorphic population, the frequency of an adaptive $\delta$-variant $\delta_i\rightarrow \delta_i(1-\epsilon)$ obeys 
\begin{equation}
\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
\end{equation}
The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare \cite[pp. 29]{crow_1970}). On the other hand, the frequency of an adaptive $b$-variant $b_i\rightarrow b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.

In practice the density dependence in Eq.~\eqref{eq:Ndependentsweep} only matters if $N$ changes substantially during a sweep. This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion). A more serious objection to Eq.~\eqref{eq:canonical} is that it is internally inconsistent even when $N$ has reached equilibrium. To quantify how serious this objection is, we need to account for how much $N$ changes as a result of a sweep involving a density-regulating trait. In Eq.~\eqref{eq:simplebirthdeath}, the saturation density is $N_{\rm initial}=b_i/\delta_i$, which increases to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$ over a $\delta$ sweep. Thus, the selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$ over a $\delta$ sweep (Fig.~\ref{fig:strengthofselection}). Consequently, Eq.~\eqref{eq:canonical} breaks down if selection is sufficiently strong on $\delta_i$, with proportional effects of $\epsilon>0.2$ inducing substantial deviations from constant selection. The cause of this effect is that the mortality rate $\delta_i N$ is both density-regulating and subject to density-dependent selection; thus selection on $\delta$ induces a change in $N$ but also changes in strength in response to $N$.

@@

It is widely recognized that the relative fitness description of selection is poorly integrated with the ecological literature on population growth \citep{mallet_2012}. This is not simply a lack of ecological realism on the part of relative fitness models. In many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}. As a result, competition and the selection it induces can concentrated at the juvenile phase, where a large reproductive excess allows for for most juveniles to be competitively eliminated. In our lottery model, this excess appears when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, artificially invalidating Eq.~\eqref{eq:canonical}. In this respect, the Wright-Fisher model and similar ``viability selection'' models actually capture an important ecological process.

Having highlighted this redeeming quality of Eq.~\eqref{eq:canonical}, we now turn to its breakdown. We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes (adult) population density and is also density-dependent. In our lottery model, this can only occur if types differ in $c$ as well as $b$ or $d$. Why does our model need joint trait variation when the classical density-dependent selection literature is built on traits that behave like the $\delta$ in Eq.~\eqref{eq:canonical} \citep{christiansen_2004}? To briefly review: Based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme posited a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Thus we are left with $r$-selection (uncrowded) versus $K$- or $\alpha$-selection (crowded). The latter two forms of selection in crowded populations behave like selection on $\delta$ in that they are both density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep).

By contrast, the density-dependent ($c$) and density-determining traits ($b$ and $d$) in our lottery model are independent. This trait separation highlights the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1976}. Both are needed to pose a threat to Eq.~\eqref{eq:canonical} (recall that we are still assuming $N$ is not far from equilibrium). In this respect the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012} actually represent a complicated form of the density-selection interaction that confounds a key underlying issue.

Relative fitness models truly break down if $N$ is far from equilibrium and selection is density-dependent. For example, wild \textit{Drosophila} experience huge seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}. In this case there is no choice but to abandon relative fitness. Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful, because drift in our model should behave broadly similarly. Thus, our model it should provide a useful starting point for analyzing evolution in this and other from-from-equilibrium situations.

>>>>

\textit{Keywords}: Lottery model, competitive Lotka-Volterra, $r$/$K$-selection, interference competition, $R^*$ theory, eco-evo.

@@

Evolutionary biologists frequently describe selection using relative fitness, and relative fitness models are foundational in population genetics. Yet the population ecological basis of relative fitness is poorly understood, and the classical ecology literature on selection in crowded populations seems to be incompatible with widely-used relative fitness models such as the Wright-Fisher. Here we develop a generalization of the Wright-Fisher model in which density depends dynamically on the demographic rates of the types present. We then explore the population ecology of relative fitness using this model as a reference point. Although density-dependent selection can cause relative fitness to break down, relative fitness models are fairly robust in populations close to demographic equilbrium. In particular, relative fitness only breaks down if selection is strong and density-dependent, and density is also selection-dependent. Our generalized Wright-Fisher model clearly distinguishes between three demographic parameters, each of whose behavior is, on its own, correctly described by relative fitness models. In contrast, the classical literature on density-dependent selection confounds them. We argue that selection-independent density is ecologically plausible for many species given the prevalence of reproductive excesses, an important aspect of selection omitted in many ecological models. Our model also offers a natural alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.

@@

There are a variety of different measures of fitness. Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, because different measures  may be more useful in different circumstances. But it should be clear how the measure being used is connected to the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}. While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness seems largely divorced from population biology. It has even been proposed that relative fitness be justified from more abstract measure-theoretical arguments, abandoning population biology altogether \citep{wagner_2010}.

In uncrowded populations, relative fitness simply represents differences in intrinsic population growth rate. In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$. Here we can rescale $W$ however we please and replace it with ``relative fitness'' $w$ without affecting the ratio $\frac{W_i}{\overline{W}}=\frac{w_i}{\overline{w}}$. In continuous time, the canonical selection equation is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W$ is replaced by the intrinsic exponential growth rate $r$ \citep[pp. 26]{crow_1970}. If there are two types present, a wildtype and a mutant for instance, then the continuous time canonical selection equation can be written as
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where the constant selection coefficient $s$ is the difference in $r$ between types. The corresponding adaptive sweeps follow a logistic curve.

@@

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. [will make this sound less dismissive] Adaptive dynamics currently provides a powerful framework for addressing the complex feedbacks between evolutionary change and population density, including but not limited to extinction \citep{diekmann_2004}. However, the focus of adaptive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''. We emphasize that our focus here is the description of the time-dependent process by which selection changes allele frequencies, which is particularly critical for making sense of evolution at the genetic level for which we now have abundant data.

Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. Rather attempting to make sense of relative fitness in existing standard models of population growth mentioned above (e.g. \citep{kimura1969natural,mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher, constant-$N$, relative fitness model. Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}. Like the Wright-Fisher model, the classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}. Unlike the Wright-Fisher model, generations can overlap in the lottery model. We generalize the lottery model to allow population density to vary, creating a density-dependent generalization of the Wright-Fisher model with overlapping generations.

We show that when this model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges. Futhermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded populations. In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities. This intuition is largely an artifact of the models historically used in the density-dependent selection literature, which ignore the existence of relative contests.

Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ). We then...[BLAH]

@@

When multiple propagules land on the same territory, the victor is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$. Type $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these. Ignoring fluctuations about these two expectations (due to our no-drift, large $T$, large $n_i$ approximation), type $i$'s territorial acquisition is given by
\begin{equation}
\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G) \label{eq:growthsumuncoupled}
\end{equation}
in our extended lottery model, where the sum only includes territories with at least one propagule present.

@@

Eq. \eqref{eq:lottery} breaks down for types with low propagule density ($l_i\ll 1$) because territorial acquisition is then not correctly represented by a lottery in each territory with the mean propagule density. Instead, the propagules from a low density type only contest a small fraction of territories, and do so with at least one propagule present (not a small fraction of a propagule $l_i\ll 1$). In section ``Analytical approximation of the density-dependent lottery'', we present a generalization of \eqref{eq:lottery} that accounts for this low density behavior.

@@

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second, $R_i c_i/\overline{c}$, represents competitive victories when the $i$ type is a rare invader in a high density population, determining its invasion fitness \citep{metz_1992}. The third term, $A_i c_i/\overline{c}$, represents competitive victories when the $i$ type is abundant. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $m_i/M$. If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$). Note that not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied; total population density thus obeys 
\begin{equation}
\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
\end{equation}

@@

We are now in a position to analyze the validity of Eq.~\eqref{eq:canonical} more explicitly. In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's general argument \citep{macarthur_1967} never precluded this possibility. Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth. To understand why this poses a problem for Eq.~\eqref{eq:canonical}, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in type $i$'s mortality rate due to crowding (for simplicity, there are no deaths when uncrowded). Then, starting from a monomorphic population, the frequency of an adaptive $\delta$-variant $\delta_i\rightarrow \delta_i(1-\epsilon)$ obeys 
\begin{equation}
\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
\end{equation}
The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare with Model III in \cite{kimura1969natural}). On the other hand, the frequency of an adaptive $b$-variant $b_i\rightarrow b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.

In practice the density dependence in Eq.~\eqref{eq:Ndependentsweep} only matters if $N$ changes substantially during a sweep. This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion). However, even if $N$ has reached equilibrium, it will change substantially over a $\delta$-sweep if selection on $\delta$ is sufficiently strong. To quantify this effect, we need to account for how much $N$ changes as a result of a $\delta$-sweep beginning and ending in equilibrium \citep{kimura1969natural}. The saturation density in Eq.~\eqref{eq:simplebirthdeath} is $N_{\rm initial}=b_i/\delta_i$, which increases to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$ over the $\delta$-sweep. Thus, the selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occur with proportional changes to $\delta$ of order $\epsilon=0.1$ and upwards (Fig.~\ref{fig:strengthofselection}).

The cause of this effect is that the mortality rate $\delta_i N$ is both density-regulating and subject to density-dependent selection; thus selection on $\delta$ induces a change in $N$ but also changes in strength in response to $N$.

@@

It is widely recognized that the relative fitness description of selection is poorly integrated with the ecological literature on population growth \citep{mallet_2012}. This is not simply a lack of ecological realism on the part of relative fitness models. Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}. As a result, competition and the selection it induces can be concentrated at the juvenile phase, where a large reproductive excess allows for most juveniles to be competitively eliminated. In our lottery model, this excess appears when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, artificially invalidating Eq.~\eqref{eq:canonical}. In this respect, the Wright-Fisher model and similar ``viability selection'' models actually capture an important ecological process.

We now turn to the breakdown of Eq.~\eqref{eq:canonical}. We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes (adult) population density and is also density-dependent. In our lottery model, this can only occur if and only if types differ in $c$ as well as $b$ or $d$. Why does our model need joint trait variation when the classical density-dependent selection literature is built on traits that behave like the $\delta$ in Eq.~\eqref{eq:canonical} \citep{christiansen_2004}? To briefly review: Based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme posited a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Thus we are left with $r$-selection (uncrowded) versus $K$- or $\alpha$-selection (crowded). The latter two forms of selection in crowded populations behave like selection on $\delta$ in that they are both density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep).

By contrast, the density-dependent ($c$) and density-determining traits ($b$ and $d$) in our lottery model are independent. This trait separation highlights the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}. The combination is needed to pose a threat to Eq.~\eqref{eq:canonical} (recall that we are still assuming $N$ is not far from equilibrium). In this respect the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012} actually represent a complicated form of the density-selection interaction that confounds a key underlying issue.

Note that while this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear how we should expect the trait variation in nature to align. For instance, should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent? In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent. However, for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it any interference competition are subsumed into the competitive ability $c$, which does not affect $N$.

Relative fitness models truly break down when $N$ is far from equilibrium and selection is density-dependent. For example, wild \textit{Drosophila} experience huge seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}. In this case there is no choice but to abandon relative fitness, and our model provides one potentially suitable option. Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful, because drift in our model should behave broadly similarly. Thus, our model it should provide a useful starting point for analyzing evolution in this and other from-from-equilibrium situations.
